---
title: "Modeling Individual and Spatial Variation with Hierarchical Models" 
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: "Modeling Individual and Spatial Variation with Hierarchical Models"
---

```{r setup, include=FALSE}
library(learnr)
#require(rstanarm)
#require(bayesplot)
require(ggplot2)
require(lhs)
require(plotly)

require(purrr)
require(faux)
require(shiny)
require(dplyr)
require(ggplot2)
require(nlme)

#### START OF multilevel.R CODE ####

## Simulate from a multilevel model

## N = Number of individuals,
## G = Number of neighborhoods


generateNeighborhoodData <- function(r, beta_n, neighborhood_sd) {
  G <- 30
  N <- 1000
  
  
  alpha <- 120
  #neighborhood_sd <- 5
  
  
  ## Sample individuals into neighborhoods w/equal probability
  df <- data.frame(
    neighborhood = 1:G,
    N = rmultinom(1:G, N, rep(1 / G, G)),
    walkability = rnorm(G, 0, 1)
  ) %>%
    mutate(
      mu_income = rnorm_pre(walkability, 0, 1, r = r),
      n_alpha = rnorm(G, alpha, neighborhood_sd) + beta_n *
        walkability
    ) %>%
    filter(N > 0)
  
  return(df)
}

generateIndividualData <- function(df, beta_income) {
  ind_sd <- 1
  
  neighborhoods <-
    pmap_dfr(df, function(neighborhood,
                          N,
                          walkability,
                          mu_income,
                          n_alpha) {
      data.frame(neighborhood = neighborhood,
                 income = rnorm(N, mu_income, 1)) %>%
        mutate(y = rnorm(N, n_alpha + beta_income * income, ind_sd))
    }) %>% inner_join(df)
  
  
  return(neighborhoods)
}

histogramUI <- function(id) {
  sidebarLayout(sidebarPanel(
    h4("Generate Neighborhood-Level Data"),
    p(
      "Use the sliders to examine different relationships between neighborhood walkability, average income, and average blood pressure:",
      style = "font-style: italics;"
    ),
    sliderInput(
      NS(id, "nsd"),
      "Standard Deviation of Neighborhood-Level Errors",
      min = 0.5,
      max = 10,
      step = 0.1,
      value = 2
    ),
    sliderInput(
      NS(id, "beta_w"),
      "Change in Average SBP for 1-unit Changed in Walkability",
      min = -4,
      max = 0,
      step = 0.1,
      value = -1
    )
  ),
  
  mainPanel(plotlyOutput(NS(id, "bp"))))
}


histogramServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    ## Generate the model data in response to parameter inputs
    neighborhoodData <-
      reactive(generateNeighborhoodData(input$r, input$beta_w, input$nsd))
    individualData <-
      reactive(generateIndividualData(neighborhoodData(), input$incomebp))
    
    output$hist <- renderPlotly({
      df <- neighborhoodData()
      g <- ggplot(df, aes(x = walkability, y = mu_income)) +
        geom_point() +
        geom_smooth(method = "lm", se = FALSE) +
        xlab("Neighborhood Walkability") +
        ylab("Neighborhood Average Income") +
        xlim(-2.5, 2.5) +
        ylim(-2.5, 2.5)
      p <- ggplotly(g)
      return(p)
    })
    
    output$bp <- renderPlotly({
      df <- neighborhoodData()
      g <- ggplot(df, aes(x = walkability, y = n_alpha)) +
        geom_point() +
        geom_smooth(method = "lm", se = FALSE) +
        xlab("Neighborhood Walkability") +
        ylab("Neighborhood Expected SBP") +
        xlim(-2.5, 2.5)
      p <- ggplotly(g)
      return(p)
    })
    
    
    output$income <- renderPlotly({
      df <- neighborhoodData()
      g <- ggplot(df, aes(x = mu_income, y = n_alpha)) +
        geom_point() +
        geom_smooth(method = "lm", se = FALSE) +
        xlab("Neighborhood Avg. Income") +
        ylab("Neighborhood Expected SBP") +
        xlim(-2.5, 2.5)
      p <- ggplotly(g)
      return(p)
    })
    
    output$indbp <- renderPlotly({
      df <- individualData()
      g <- ggplot(df, aes(x = income, y = y)) +
        geom_point() +
        geom_point(aes(colour = walkability)) +
        geom_smooth(method = "lm", se = FALSE) +
        xlab("Individual Income") +
        ylab("Individual SBP") +
        xlim(-2.5, 2.5)
      p <- ggplotly(g)
      return(p)
    })
    
    simpleModel <- reactive({
      if (input$adjust == TRUE) {
        f <- y ~ income + walkability
      } else {
        f <- y ~ income
      }
      
      return(lm(f, data = individualData()))
    })
    
    multilevelModel <- reactive({
      if (input$adjust == TRUE) {
        f <- y ~ income + walkability
      } else {
        f <- y ~ income
      }
      return(nlme::lme(f, random = ~ 1 |
                         neighborhood, data = individualData()))
      
    })
    
    selectedModel <- reactive({
      if (input$hierarchical == TRUE) {
        return(multilevelModel())
      } else {
        return(simpleModel())
      }
    })
    
    output$modelOut <- renderPrint({
      return(summary(selectedModel()))
    })
    
    output$modelResid <- renderPlotly({
      m <- selectedModel()
      df <- data.frame(y = resid(m))
      g <- ggplot(df, aes(sample = y)) + stat_qq() + stat_qq_line()
      p <- ggplotly(g)
      return(p)
      
    })
    
    output$incomeResid <- renderPlotly({
      df <- individualData()
      x <- df$income
      df$resid <- resid(selectedModel())
      
      g <-
        ggplot(df, aes(x = income, y = resid)) + geom_point(aes(colour = walkability)) + geom_smooth(method =
                                                                                                       "lm", se = FALSE)
      p <- ggplotly(g)
      return(p)
    })
    
    output$beta_income_val <- renderText({
      return(paste0(
        "Input Value of Income Effect = ",
        as.character(input$incomebp)
      ))
    })
    
    output$beta_walk_val <- renderText({
      return(paste0(
        "Input Value of Walkability Effect = ",
        as.character(input$beta_w)
      ))
    })
  })
  
}

individualUI <- function(id) {
  sidebarLayout(sidebarPanel(
    h4("Generate Individual-Level Data"),
    p(
      "Manipulate the relationship between individual-level income and individual SBP.",
      style = "font-style: italics;"
    ),
    sliderInput(
      NS(id, "incomebp"),
      "Change in SBP for Each 1-unit Change in Income",
      min = -4,
      max = 0,
      step = 0.1,
      value = -1
    )
  ),
  mainPanel(plotlyOutput(NS(id, "indbp")),))
}

modelUI <- function(id) {
  sidebarLayout(
    sidebarPanel(
      h4("Analyze Individual Level Data"),
      checkboxInput(NS(id, "adjust"), "Adjust for Walkability", value = FALSE),
      checkboxInput(NS(id, "hierarchical"), "Use Hierarchical Model", value = FALSE),
      h4("Input Parameter Values"),
      textOutput(NS(id, "beta_income_val")),
      textOutput(NS(id, "beta_walk_val"))
      
      
      
    ),
    mainPanel(tabsetPanel(
      tabPanel("Model", verbatimTextOutput(NS(id, "modelOut"))),
      tabPanel("Normality", plotlyOutput(NS(id, "modelResid"))),
      tabPanel("Residuals", plotlyOutput(NS(id, "incomeResid")))
    ))
  )
  
}

correlationUI <- function(id) {
  sidebarLayout(sidebarPanel(
    sliderInput(
      NS(id, "r"),
      "Correlation Between Neighborhood Walkability and Average Income",
      min = 0,
      max = 1,
      step = 0.02,
      value = 0.0
    ),
  ),
  mainPanel(plotlyOutput(NS(id, "hist"))))
  
}


histogramApp <- function() {
  ui <- fluidPage(
    histogramUI("hist1"),
    individualUI("hist1"),
    modelUI("hist1"),
    correlationUI("hist1")
  )
  server <- function(input, output, session) {
    histogramServer("hist1")
  }
  shinyApp(ui, server)
}
#### END OF multilevel.R CODE ####
knitr::opts_chunk$set(echo = FALSE)
```

## Learning Objectives

The point of this interactive exercise is to introduce you to several issues in the analysis of multi-level, nested data, and the ways tools like hierarchical models can - and can't - help you address them. Specifically, we'll focus on: 

1. **Identifying** sources of individual- and neighborhood-level variability in a dataset consisting of individual-level observations;
2. **Examining** how hierarchical modeling approaches can adjust for unmeasured neighborhood-level confounders; and
3. **Understanding** the potential for mis-estimation of key effects under different balances of **systematic** and **random** variation at the neighborhood level.

## Thinking Hypothetically
### Neighborhoods, Wealth, and Blood Pressure

In this example, imagine individuals are grouped into discrete neighborhoods by wealth (not an unrealistic assumption), and that increasing wealth is also protective against SBP at the individual level.

Now, we'll let a factor that is potentially associated with SBP risk (such as walkability, which could decrease SBP or exposure to pollution which could increase it) vary across neighborhoods and look at what happens when we let that factor increase or decrease along with average neighborhood wealth. 

Specifically, we're looking at a kind of classic **multilevel** or **hierarchical** model. Again we're interested in a **continuous** outcome at the individual level, i.e. systolic blood pressure (SBP). 

We're going to model variation in this outcome as a function of:


1. An **individual factor** - in this case wealth - represented by the **covariate** $x_i$ and model coefficient $\beta$. 
    - We assume that wealth is protective, i.e. as wealth goes up, SBP goes down. So we constrain the value of $\beta \le 0$. 

2. An **unobserved attribute** of each neighborhood that is associated with either increased or decreased SBP, denoted by the coefficient $\gamma$, which can take on both positive and negative values. 
    - To keep things simple, we assume that the intensity of this effect increases from neighborhood 1 to neighborhood 6, just to make things easier to visualize.

### A Simple Model of Individual and Contextual Risk

In this model we're making two more simplifying assumptions which we can relax later on:

1. Individuals sort **deterministically** into neighborhoods by wealth. The population is just sliced up into 6 neighborhoods with the poorest 1/6 of the population in neighborhood 1, and the wealthiest in neighborhood 6, etc. 

2. The **effects** of living in each neighborhood are also deterministic, i.e. there is no randomness associated with the effect of living in Neighborhood 3 vs. Neighborhood 1 in SBP. 

When we put this all together, we have a model that looks like this:

$y_{ij} = \alpha + \beta x_i + \gamma z_j + \epsilon_i$

In other words, the expected value of SBP for individual $i$ in neighborhood $j$ is a function of their wealth ($x_i$), a condition in their neighborhood of residence ($z_j$) and normally distributed variaton at the individual level ($\epsilon_i$), where $\epsilon_i \sim \text{Normal}(0, \sigma)$ where $\sigma$ is the standard devation of individual SBP outcomes.

Ok, now that we've laid out some of the preliminaries, time to see what this looks like!

## Generate Some Data
### Change in SBP for Each 1-Unit Change in Individual Wealth
Move the slider to create an effect in which increasing wealth is associated with decreasing systolic blood pressure, i.e. for each 1-unit increase, there's a **decrease** in SBP.

```{r}
sidebarPanel(
  sliderInput(
    "beta_1_n",
    "Change in SBP for Each 1-Unit Change in Individual Wealth (smaller values indicate more protective effect of wealth):",
    min = -3,
    max = 0,
    value = 0
  ),
)

mainPanel(plotlyOutput("neighborhoodPlot"))


```

The blue line overlaid on the plot is a regression line showing the association between wealth and SBP estimated from these data.

### Including an **Unobserved** Component of Neighborhood-Level Variation

The data above reflect a relationship in which individual wealth is **protective** against cardiovascular disease - at least as measured by systolic blood pressure. 

Now, we can bring in a piece of neighborhood context that is shared by all members of each neighborhood: 

Let's imagine that neighborhoods 1-5 sit on a continuum from least-to-most walkable as well, and that this has an independent protective impact on the health of all individuals in each neighborhood, regardless of their individual wealth. 

You can manipulate the slider below to see how the visual and quantitative association between wealth and SBP changes when we introduce a decrease in SBP with increasing walkability, which is itself correlated with average neighborhood-level wealth.

```{r}
sidebarPanel(
  sliderInput(
    "beta_2_n",
    "Difference in Average SBP Between Least and Most-Walkable Neighborhood:",
    min = -10,
    max = 10,
    value = 0
  ),
  checkboxInput("n_adjust", "Adjust for Neighborhood", FALSE),
  checkboxInput("w_adjust", "Adjust for Walkability", FALSE)
  
)

mainPanel(
  plotlyOutput("neighborhoodPlot2")
  )
fluidRow(
  column(12,
           verbatimTextOutput("n_model_summary")
         )
)

```

#### Additional things to try

1. What happens to the estimate of the wealth effect if you adjust for walkability?
2. What happens to the wealth effect estimate if you don't adjust for walkability but **do** adjust for neighborhood?
3. What about when you adjust for wealth, walkability & neighborhood? 

Finally, tick the box to adjust for neighborhood and see how it impacts the parameter values. This model is just a simple example that treats each neighborhood as its own **categorical variable** since there are only a few of them. But this model has most of the ingredients of a classic hierarchical model. 

## More Neighborhoods & Variability

Now, we'll take a look at what this problem looks like in a slightly more complex context where:

1. We allow there to be more **random variation** in the effect of living in a given place;

2. There are **more than just a few units**; and

3. The number of individuals observed in each neighborhood is **not as large** as before.

### A New Model

In this example, we consider observations from 1000 people sampled randomly from 30 neighborhoods, more than can be easily dealt with using the approach from the previous example.

Just like in the previous example, we'll model **individual-level outcomes as a function of systematic and random variation**. In this case, systematic variation will be represented by the coefficient $\beta$, which represents the relationship between individual wealth and SBP, and normally-distributed random variation by the error term $\epsilon_i$. In other words, $\epsilon_i \sim \text{Normal}(0, \sigma_i)$, where $\sigma_i$ is the standard deviation of SBP values **between individuals**.

In addition to this, we're going to allow each neighborhood $j$ to have a mean SBP that **deviates from the population mean** SBP. This is represented by a neighborhood-level error term, which we'll call $\epsilon_j$. This value represents how far the mean value of the systolic blood pressure (SBP) outcome for individuals living in neighborhood $j$ is from the population intercept $\alpha$. 

Similarly to the individual-level error term, this is normally distributed, i.e. $\epsilon_j \sim \text{Normal}(0, \sigma_j)$, where $\epsilon_j$ is the **deviation of the mean SBP** in neighborhood $j$ from the average neighborhood-level mean across all neighborhoods. In other words, this represents **between-neighborhood variation**.

### Generate neighborhood-level mean SBP values

The first thing we're going to do is to simulate values of the neighborhood-level mean SBP. There are two components underlying variability in this example:

1. **Systematic variation** due to difference in the walkability of each neighborhood. This represents the idea that people living in neighborhoods that are more walkable may get more hours of exercise each week, contributing to their overall health and lowering the average SBP of people living in that neighborhood. In this example, walkability is represented as a **z-score**, i.e. in terms of standard deviations above or below the mean walkability of all neighborhoods.

2. **Random variation** due to unobserved neighborhood-level factors. This is what causes the neighborhood mean SBP to be larger or smaller than the mean value predicted by our systematic model. When $\sigma_N$ — the standard deviation of neighborhood-level means — is small, most of the variability between neighborhoods is due to **systematic**, observed factors.

### Generate Individual-Level SBP Values 
```{r}
histogramUI("hist1")
individualUI("hist1")
```

Now that we have simulated a set of neighborhood level means, let's simulate the individual-level data that constitute the measurements from our simulated study.

In addition to neighborhood features like walkability, other individual-level attributes are likely to impact SBP. For example, people with higher incomes may be more able to access quality care, avoid stressful work environments, etc., which may make them more likely to have lower SBP. 

To begin, we'll assume that there is **no relationship** between individual income and neighborhood walkability, i.e. a neighborhood with low average income is just as likely to be walkable as one with a high average income. Here, income values are expressed as a z-score, with values indicating standard deviations above or below the mean income across all neighborhoods.

Move the **"Change in SBP for Each 1-unit Change in Income" slider** to vary the protective effect of income on SBP. Here, the colors of the points represent the walkability of the neighborhoods that individuals live in.

## The Impact of Omitting Variables

In this section, we'll look again at the impact of **omitting variables** from our regression model on the conclusions we get from it. 

### Estimating the Effects of Wealth, Walkabaility, and Neighborhood on Simulated SBP

Here, imagine we are conducting a study looking just at the relationship between individual income and SBP, and we have no information on the individual neighborhoods people live in. Take a look at what happens when you estimate the effect of income on SBP with and without walkability in the model, and using a regular linear regression model (the `lm` function in R) or a hierarchical regression model that accounts for **neighborhood-level clustering** of risk. 

```{r}
modelUI("hist1")
```

Take a look at the different panels, which show
1. The model results, 
2. The distribution of the model **residuals** as compared to a normal distribution (known as a [Q-Q plot](https://en.wikipedia.org/wiki/Q%E2%80%93Q_plot)), and 
3. A plot of the model residuals plotted against our variable of interest, income. 

#### Additional Questions

1. Under what conditions do you get an over-estimate of the protective impact of income on SBP?
2. Does adjusting for walkability without accounting for neighborhood **clustering** get better results? When does it not work to do this?
3. Are there situations in which adjusting for neighborhood clustering does not bring about the appropriate values of the input values?

### Introducing correlation between **observed** and **unobserved** factors.

Now, we're going to try to see what happens when we add some more complexity. Up until this point, we've assumed that neighborhood wealth and walkability are unrelated. Obviously this doesn't square with our experiences of the real world, in which certain desirable attributes, such as walkability, lead to an increase in prices and the average wealth of people living there.

Use the slider below to change this relationship by increasing the correlation between neighborhood level average income and and neighborhood level walkability. Also go back up to the other sections and try to change other input settings and see what factors impact your ability to get accurate estimates and fulfill the requirements of a linear regression model (i.e. normal, uncorrelated residuals).

```{r}
correlationUI("hist1")
```

```{r, context="server"}
histogramServer("hist1")
```

```{r, context = "server"}
group_indices <- function(N, J) {
  out_groups <- c()
  split_size <- N %/% J
  
  for (i in 1:(J - 1)) {
    out_groups <- c(out_groups, rep(i, split_size))
  }
  
  if ((N %% split_size) > 0) {
    last_indices <- rep(J, N %% split_size)
  } else {
    last_indices <- rep(J, split_size)
  }
  out_groups <- c(out_groups, last_indices)
  
  return(out_groups)
}



area_data <- reactive({
  J <- 5
  N <- 1000
  alpha_n <- 140
  
  xvals <- 10 * sort(as.vector(randomLHS(N, 1)))
  group_ids <- group_indices(N, J)
  walkability <- (group_ids - 1) / (J - 1)
  
  y_hat <-
    alpha_n + input$beta_1_n * xvals + input$beta_2_n * walkability
  
  y_sim <- rnorm(length(y_hat), y_hat, 2)
  
  df <-
    data.frame(
      x = xvals,
      walkability = walkability,
      neighborhood = as.factor(group_ids),
      y = y_sim
    )
  
  return(df)
  
})

neighborhoodPlot <- renderPlotly({
  g <- ggplot(area_data(), aes(x = x, y = y)) +
    geom_point(aes(colour = neighborhood)) +
    xlab("Wealth") +
    ylab("SBP")
  
  if (input$n_adjust == TRUE) {
    g <-
      g + geom_smooth(method = "lm",
                      aes(group = neighborhood),
                      se = FALSE)
  } else {
    g <- g + geom_smooth(method = "lm", se = FALSE)
  }
  
  p <- ggplotly(g)
  return(p)
})
output$neighborhoodPlot <- neighborhoodPlot
output$neighborhoodPlot2 <- neighborhoodPlot

n_model_fit <- reactive({
  if (input$n_adjust == TRUE) {
    if (input$w_adjust == TRUE) {
      fit <- lm(y ~ x + walkability + neighborhood, data = area_data())
      names(fit$coefficients) <-
        c(
          "Intercept",
          "Wealth",
          "Walkability",
          "Neighborhood 2",
          "Neighborhood 3",
          "Neighborhood 4",
          "Neighborhood 5"
        )
    } else {
      fit <- lm(y ~ x + neighborhood , data = area_data())
      names(fit$coefficients) <-
        c(
          "Intercept",
          "Wealth",
          "Neighborhood 2",
          "Neighborhood 3",
          "Neighborhood 4",
          "Neighborhood 5"
        )
    }
    
  } else {
    if (input$w_adjust == TRUE) {
      fit <- lm(y ~ x + walkability, data = area_data())
      names(fit$coefficients) <-
        c("Intercept", "Wealth", "Walkability")
    } else {
      fit <- lm(y ~ x, data = area_data())
      names(fit$coefficients) <- c("Intercept", "Wealth")
    }
    
    
  }
  return(fit)
})

output$n_model_summary <- renderPrint({
  summary(n_model_fit())
})
```